---
title: "Konza Vegetation Analysis"
author: "Angelina Kirton"
format: html
---

## Introduction

This analysis will look at vegetation data from the Konza National Ecological Observatory Network (NEON) site located near Manhattan, KS. It will use tidyverse packages to wrangle and visualize some of the data.

## Loading the data 

The data are provided in csv.files. We will use the tidyverse package 'readr' to load it into R.

```{r}
#| label: read-csv

# load packages
library("readr")
library("dplyr")
library("ggplot2")

# read in the data
konza_data <- read_csv("data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z (1).csv")
```

## Data summarizing with 'dplyr'

We will next use the dplyr package to summarize some different aspects of the data.

```{r}
#| label: summarize-with-dplyr

# first summary to make will be a count of the number of observations in each family.
konza_data %>%
  group_by(family) %>%
  tally() %>%
  arrange(desc(n))
```

This analysis shows that Poaceae and Asteraceae are among the most common family found in these plots.

```{r}
#| label: mean-pcnt-cover-by-family

# the goal is to calculate the mean percent cover of each family
konza_data %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover ))
```

This analysis showed that the largest percent cover is the 'NA' taxa, which means it likely was not able to be identified, but the highest mean percent cover at 16% across the plots was Fagaceae.

```{r}
#| label: filter-by-elevation

# the goal is to calculate the mean percent cover of each family
konza_data %>%
  filter(elevation > 390) %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
  arrange(desc(mean_pcnt_cover ))
```

Next we will learn to use 'mutate()' which is a function to create a new column based on contents from an existing column.

```{r}
#| label: use-mutate

# select pulls out the column you want 
# mutate creates a new column with whatever content you want
# which can include manipulation of an existing column

konza_data %>%
  select(scientificName, family, percentCover) %>%
  mutate(propCover = percentCover / 100)
```


